Value-Suppressing Uncertainty Maps

### 2017-03-31 paper/template.tex
Use figrefs
Typos and violin plot citation

There are a number of techniques for communicating values and uncertainty in one-dimensional statistical graphics~\cite{olston2002visualizing, wickham201140, neyman1937outline}. Techniques such as error bars, confidence bands, and violin plots use a a separate position channel to encode uncertainty. The techniques in this paper also apply to 2D visualizations such as heatmaps. Today, there are no established standards of visualizing uncertainty in heatmaps.

@@

\emph{Juxtaposition} creates two explicit visualizations: an ``uncertainty map'' and a ``data map.'' MacEachren~\cite{maceachren1992visualizing} portrays uncertainty visualization as the design problem of how to \emph{unify} these two maps, and so this explicit separation can be counterproductive to the goal of information fusion. Juxtaposition also creates an additional search task for viewers, who must register hot spots in one map with corresponding features of the other map. Nevertheless, especially in cases where the quantification of uncertainty is a separate analysis process from the measuring of data values, juxtaposition is a common design tactic for conveying uncertainty~\cite{moritz2017trust}.

@@

In \figref{fig:viral2d} and \figref{fig:viralVsum}, we visualized the variability in viral genomes across 25 populations of the simian immunodeficiency virus (SIV)~\cite{o2012conditional}. Next generation sequencing (NGS) allows biologists to collect and analyze large amounts of genomics data. However, these techniques often create data quality issues, as they must align large numbers of potentially ambiguous ``reads'' of base pairs. In the dataset here, it is important not only to identify hotspots in the viral genome (locations with high rates of mutation or variability), but to not be distracted by false positives. Extending the ``confidence fog'' metaphor used in viral genome browers like LayerCake~\cite{correll2015layercake}, VSUMs enable biologists to explore interesting patterns while not being distracted by noise, as shown in \figref{fig:viralVsum}.

@@

For each task, participants saw one of three types of thematic maps. Either two \textbf{juxtaposed} maps side-by-side, with separate encodings for value and uncertainty, a traditional \textbf{bivariate} map, where value and uncertainty are orthogonally encoded, or a \textbf{VSUM}. Figure \ref{fig:conditions} shows examples of these factor levels. Other factors included the size of the map (either a 4x4 or 8x8 grid), and (for the prediction task), the question framing (attack or defense). All factors were presented within subjects. Secondary factors were evenly blocked across stimuli. Fatigue effects were significant in piloting, and so we limited the number of replicates: participants saw 18 stimuli (6 for each map type) for the identification task, and 12 (4 for each map type) for the prediction task.

In all cases, we generated color maps based on the requirement that no two colors in the resulting color scheme were closer than 18 units of Eulcidean distance in CIELAB space. This threshold results in a $9$ color bivariate map, $15$ color VSUM for all color scales used in this study. Given the size of the marks used in our study, this threshold distance is sufficient for colors to be correctly distinguished in over 95\% of trials~\cite{stone2014engineering}. Other choices of threshold, and differing univariate scales, would result in different maps (see Fig.~\ref{fig:performance}).

@@

We included both a \textbf{gain} framing (given predictions of where your opponent's ships are, where should you launch your missiles?), and a \textbf{loss} framing (given predictions of where your opponent's missiles will fall, where should you place your ships?) in order to account for systematic differences in human judgments under uncertainty. Tversky \& Kahneman~\cite{tversky1985framing} illustrate that framings in terms of gains or losses produce reliably different outcomes. In general, the function of the value of gains is concave, but is convex for losses~\cite{kahneman1979prospect}. That is, avoiding a large loss is often more valuable than striving for a similarly large gain. To avoid potential (in this case, incorrect) transfer learning, the attacking and defending used different value color maps (Plasma and Cool) than the identification tasks, and from each other.  Figure \ref{fig:taskTwoConditions} shows these differing color maps and framings.

@@

As with the identification task, a one-way ANOVA also indicated that chart type had a significant effect on response time ($F(2,861)=13$, $p<0.001$). A Tukey's HSD confirmed that, as with the previous task, response time was significantly longer with juxtaposed maps (trimmed mean response time of $23$ seconds) compared to the other chart types (both with trimmed mean response time of $18$ seconds). Despite the documented effect of framing in prior work, a one-way ANOVA of task frame on expected value found no significant main effect ($F(1,862)=0.01$, $p=0.9$), and the trimmed means were quite similar (expected value of $0.76$ per guess for attackers, verus $0.75$ per guess for defenders).

>>>>

There are a number of techniques for communicating values and uncertainty in one-dimensional statistical graphics~\cite{olston2002visualizing, wickham201140, neyman1937outline}. Techniques such as error bars, confidence bands, and violin plots~\cite{hintze1998violin} use a separate position channel to encode uncertainty. The techniques in this paper also apply to 2D visualizations such as heatmaps. Today, there are no established standards of visualizing uncertainty in heatmaps.

@@

\emph{Juxtaposition} creates two explicit visualizations: an ``uncertainty map'' and a ``data map.'' MacEachren~\cite{maceachren1992visualizing} portrays uncertainty visualization as the design problem of how to \emph{unify} these two maps, and so this explicit separation can be counterproductive to the goal of information fusion. Juxtaposition also creates an additional search task for viewers, who must register hotspots in one map with corresponding features of the other map. Nevertheless, especially in cases where the quantification of uncertainty is a separate analysis process from the measuring of data values, juxtaposition is a common design tactic for conveying uncertainty~\cite{moritz2017trust}.

@@

In \figref{fig:viral2d} and \figref{fig:viralVsum}, we visualized the variability in viral genomes across 25 populations of the simian immunodeficiency virus (SIV)~\cite{o2012conditional}. Next generation sequencing (NGS) allows biologists to collect and analyze large amounts of genomics data. However, these techniques often create data quality issues, as they must align large numbers of potentially ambiguous ``reads'' of base pairs. In the dataset here, it is important not only to identify hotspots in the viral genome (locations with high rates of mutation or variability), but to not be distracted by false positives. Extending the ``confidence fog'' metaphor used in viral genome browsers like LayerCake~\cite{correll2015layercake}, VSUMs enable biologists to explore interesting patterns while not being distracted by noise, as shown in \figref{fig:viralVsum}.

@@

For each task, participants saw one of three types of thematic maps. Either two \textbf{juxtaposed} maps side-by-side, with separate encodings for value and uncertainty, a traditional \textbf{bivariate} map, where value and uncertainty are orthogonally encoded, or a \textbf{VSUM}. \figref{fig:conditions} shows examples of these factor levels. Other factors included the size of the map (either a 4x4 or 8x8 grid), and (for the prediction task), the question framing (attack or defense). All factors were presented within subjects. Secondary factors were evenly blocked across stimuli. Fatigue effects were significant in piloting, and so we limited the number of replicates: participants saw 18 stimuli (6 for each map type) for the identification task, and 12 (4 for each map type) for the prediction task.

In all cases, we generated color maps based on the requirement that no two colors in the resulting color scheme were closer than 18 units of Euclidean distance in CIELAB space. This threshold results in a $9$ color bivariate map, $15$ color VSUM for all color scales used in this study. Given the size of the marks used in our study, this threshold distance is sufficient for colors to be correctly distinguished in over 95\% of trials~\cite{stone2014engineering}. Other choices of threshold, and differing univariate scales, would result in different maps (see Fig.~\ref{fig:performance}).

@@

We included both a \textbf{gain} framing (given predictions of where your opponent's ships are, where should you launch your missiles?), and a \textbf{loss} framing (given predictions of where your opponent's missiles will fall, where should you place your ships?) in order to account for systematic differences in human judgments under uncertainty. Tversky \& Kahneman~\cite{tversky1985framing} illustrate that framings in terms of gains or losses produce reliably different outcomes. In general, the function of the value of gains is concave, but is convex for losses~\cite{kahneman1979prospect}. That is, avoiding a large loss is often more valuable than striving for a similarly large gain. To avoid potential (in this case, incorrect) transfer learning, the attacking and defending used different value color maps (Plasma and Cool) than the identification tasks, and from each other.  \figref{fig:taskTwoConditions} shows these differing color maps and framings.

@@

As with the identification task, a one-way ANOVA also indicated that chart type had a significant effect on response time ($F(2,861)=13$, $p<0.001$). A Tukey's HSD confirmed that, as with the previous task, response time was significantly longer with juxtaposed maps (trimmed mean response time of $23$ seconds) compared to the other chart types (both with trimmed mean response time of $18$ seconds). Despite the documented effect of framing in prior work, a one-way ANOVA of task frame on expected value found no significant main effect ($F(1,862)=0.01$, $p=0.9$), and the trimmed means were quite similar (expected value of $0.76$ per guess for attackers, versus $0.75$ per guess for defenders).
